From noise to knowledge: how randomness generates novel phenomena and information

### 2017-10-26 appendix/appendix.Rmd
WIP, mostly adding gillespie simulation

```{r}
library(nimble)
library(hrbrthemes)
library(ggthemes)
library(tidyverse)
ggplot2::theme_set(hrbrthemes::theme_ipsum())
```

>>>>

```{r}
library(nimble)
library(RcppRoll)
library(regimeshifts) ## devtools::install_github("cboettig/regimeshifts")
library(adaptivetau)
library(tidyverse)

## plot themes, non-essential
library(hrbrthemes)
library(ggthemes)
ggplot2::theme_set(hrbrthemes::theme_ipsum())
```

# Gillespie algorithm

Levins' Patch model as an individual-based birth-death process:

$$\frac{dn}{dt} = \underbrace{c n \left(1 - n/N \right)}_{\textrm{birth rate, } b(n)} - \underbrace{en}_{\textrm{death rate, } d(n)} $$

We simulate this using the Gillespie's exact stochastic simulation algorithm (SSA), as implemented in the `adaptivetau` R package.  We declare the transition events: birth increases the state $n$ to $n+1$, and death $n$ to $n-1$, and describe the rates associated with them:

```{r}
# indicate state variable and the change
transitions = list(c(n = +1), # birth event
                   c(n = -1)) # death event

rateF <- function(state, params, t) {
  c(params$c * state[["n"]] * (1 - state[["n"]] / params$N), # birth rate
    params$e * state[["n"]])                                 # death rate
}

We compare the stochastic simulation with identical parameters, $c=1$, $e=0.2$ except for the total number of available sites.  In our "small" system scenario, we have a total of $N=100$ sites, while in the "large" system we have $N=1000$ sites.

The theory predicts an equilibrium population size $\bar n$ at $b(\hat n) = d(\hat n)$, or

$$\left(1 - \tfrac{e}{c}\right) N,$$ 
and a variance of

$$\left. \frac{b(n) + d (n)}{2 d'(n) - b'(n)} \right\rvert_{n = \hat n} = \tfrac{e}{c}N$$

```{r gillespie}
set.seed(1234) # set random number generator seed to be reproducible

e <- 0.2
c <- 1

sim_small <- adaptivetau::ssa.exact(init.values = c(n = 50),
                       transitions, rateF,
                       params = list(c=c, e=e, N=100),
                       tf=30) %>% as_tibble()

sim_large <- adaptivetau::ssa.exact(init.values = c(n = 500),
                       transitions, rateF,
                       params = list(c=c, e=e, N=1000),
                       tf=30) %>% as_tibble()

theory <- 
list(small = 100,large = 1000) %>%
  map_dfr(function(N){
    data.frame(mean = N * (1-e/c)) %>%
    mutate(plus_sd = mean + sqrt(N * e / c),
           minus_sd = mean - sqrt(N * e / c))
  }, .id = "system_size")

gillespie <- bind_rows("large"=sim_large, 
                       "small"=sim_small,  
                       .id = "system_size") %>% bind_rows(theory)
write_csv(gillespie, "gillespie.csv")

read_csv("gillespie.csv", col_types = "cdiddd")  %>% 
  ggplot(aes(x = time)) + 
  geom_hline(aes(yintercept = mean), lty=2) + 
  geom_hline(aes(yintercept = minus_sd), lty=2) + 
  geom_hline(aes(yintercept = plus_sd), lty=2) + 
  geom_line(aes(y = n), col="red") +
  facet_wrap(~system_size, scales = "free_y") + 
  labs(caption = "theoretical predictions for mean and plus/minus one standard \n deviation shown in horizontal black dashed lines.")
```

### 2017-10-26 paper.Rmd
WIP, mostly adding gillespie simulation

<!-- Master equation transform-->

<!-- Gillespie algorithm; Population level (SDE) vs agent-based vs individual  -->

![Figure 1: demographic stochasticity arises from discrete transitions  between states in in a Markov process.  Here, a simple birth-death process has transition rates determined by the current population size, _n_.](markov.png)

>>>>

<!-- Master equation solution of birth-death model, from appendix-->

<!-- Gillespie algorithm; Population level (SDE) vs agent-based vs individual. Show figure with small and large N  -->

```{r warning=FALSE}
read_csv("appendix/gillespie.csv", col_types = "cdiddd")  %>% 
  ggplot(aes(x = time)) + 
  geom_hline(aes(yintercept = mean), lty=2) + 
  geom_hline(aes(yintercept = minus_sd), lty=2) + 
  geom_hline(aes(yintercept = plus_sd), lty=2) + 
  geom_line(aes(y = n), col="red") +
  facet_wrap(~system_size, scales = "free_y") + 
  labs(caption = 
"Population dynamics from a Gillespie simulation of the Levins model with large (N=1000) 
and small (N=100) number of sites (red) show relatively smaller effects of demographic 
noise in the larger systme. Theoretical predictions for mean and plus/minus one standard
deviation shown in horizontal black dashed lines.")
```
